Meaningful data sharing

### 2020-09-23 new-paper.Rmd
Finished rest of Karthik's tasks

The idea of open and transparent science dates back to the founding principles of The Royal Society of London in 1660. The movement, which has seen shifting support over the years, has enjoyed growing popularity over the past decade [@Nielsen2020; @Peng2011] mainly among early career researchers [@McKiernan2016]. A key motivation for practicing openness is the notion that sharing all research artefacts will allow others to reproduce and verify results, thereby improving trust and verification of those findings. Jon Clarebout famously wrote in 1992 that reproducibility in computationally oriented fields had become so trivial that non experts could do the task. Clarebout's optimistic vision was a little ahead of its time as the tools and technologies available in the early nineties for literate programming presented too high of a barrier for most researchers to adopt without incurring a significant loss in productivity [@Barnes2010].

@@

```{r pyramid, fig.cap = "Strategy for culture and behavior change, taken from the Center for Open Science: https://www.cos.io/blog/strategy-for-culture-change . To reach the top, all the lower levels need to be adequately achieved."}
# pyramid image
knitr::include_graphics("figures/pyramid.png")
```

```{r incentable-table, fig.cap = "Table of the mechanism for behavior change, the incentive, and our assessment of where the elements of Data, Code, and Computational Environment rank in terms of completing these aspects. We note that Data is often required, but the preceding steps are not required. This is in contrast to Code having no policy."}
knitr::include_graphics("figures/gt-incentive-table.png")
```

@@

Use a README file with your data project to communicate its purpose and contents. A README is often the first place users will look to learn more about a project [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards. For example, the data from Dryad, https://datadryad.org/stash/dataset/doi:10.5061/dryad.rn6k1 could be greatly enhanced by providing a README file that explains what the variables in the data collected are, and what the different sheets of the excel worksheet contain.

@@

Provide a license with your data. A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. Use the CC0 license, as it provides simple rules for attribution, warranty, and use, and is very common. The CC0 is a "public domain" license, meaning data owners waive all rights, and the work is now "owned" by the public. CC0 data can be freely shared, copied, modified, and distributed, even for commercial purposes **without asking permission**. When using data with CC0, it is good practice to cite the original work, but is not required. For more information on the benefit of using CC0, and why CCBY is not always appropriate, see https://osc.universityofcalifornia.edu/2016/09/cc-by-and-data-not-always-a-good-fit/. Short and full overviews of CCBY and CC0 can be found at [@cc0-short; @cc0-long].

Avoid using non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

@@

Make raw and analysis ready data available. Raw data is the first format of data provided before any cleaning has taken place (e.g., data in some binary or proprietary format). It should be placed in its own folder. Data used in analysis should be provided, especially if computationally intense to derive. It should be in its own folder, and to maximise interoperability, recorded in a plain text (e.g., csv, tsv) and if possible a binary format (e.g., .rds, .sav, .xlsx, Apache Parquet). The binary format allows for immediate reuse in the intended state, while the interoperability of plain text versions remain stable over time. Ideally data should be in "Tidy Data" form, where variables are in columns, observations in rows, and each cell has only one value [@Wickham2014].

@@

Share code used to transform data from its raw format into analysis ready form. This should be kept in the same place as raw data. Ideally, this transformation should involve only scripted languages (e.g, R, Python, Matlab, etc), but any non-scripted data preparation steps should be recorded in a plain text file. Authors should run the scripts in a clean environment to ensure everything runs and that the code doesn't rely on other local artefacts, such as their own folder layout not available to others.

@@

Describe how you want your data to be cited in a CITATION file. Citation is important for provenance, and credit. Citing data clarifies the data version used (provenance), and provides authors a citation (credit). To ensure these are provided, data needs to have an accession number, such as a Digital Object Identifier (DOI). A DOI is a permanent, unique identifier for a digital object such as a paper, poster, or software. Access numbers are a prerequisite for citation - if they are unavailable, a citation will be meaningless, as it cannot be tracked by any means. A file named CITATION should be placed in the directory at the top level. It should contain an accession number, and could be in `.bibtex` format. Avenues that provide DOIs are discussed in Section \@ref(avenues).

Repositories such as Dryad will manage citation. However it is worth noting that a CITATION file allows you to customise how you want your data to be cited. For example, to cite the paper that used or described the data and collection methods.

@@

We'd like to thank the following people for helpful discussions and contributions to the paper as we were preparing it for publication: Anna Krystalli, Daniella Lowenberg, and Miles McBain. We'd also like to thank those who provided feedback on the paper when it went up for presubmission: Adrian Barnett, Raymond B Huey, Christian Thurn, Christina Maimone, and GitHub user, 'dschief001'. Finally, we thank the following organisations for their support ACEMS International Mobility Programme, the Helmsley Charitable Trust, the Gordon and Betty Moore Foundation, and the Sloan Foundation.

@@

Without data, there is no science. Science needs to be reproducible so we can trust the results, and progress as a field. But without data, we do not have true reproducibility. Currently, the role of data in reproducibility is not adequately discussed. We believe this is in part because discussions around data are centered around both data quality and data sharing. However, these are in fact two separate (but related) topics. We provide immediately usable guidelines to sharing data alongside your research, with minimal barriers, that a researcher could pick up and use today.

@@

Karthik is a senior data scientist at the Berkeley Institute of Data Science and a co-founder of the rOpenSci project. Karthik is also a senior PI at the Berkeley Initiative for Global Change Biology. Prior to joining Berkeley, he earned his PhD in Ecology & Evolution from the University of California, Davis. Karthik's interests are focused on reproducible research, especially as it applies to global change. Much of his recent work focuses on building tools and services around open data and growing diverse data science communities.

>>>>

The idea of open and transparent science dates back to the founding principles of The Royal Society of London in 1660. The movement, which has seen shifting support over the years, has enjoyed growing popularity over the past decade [@Nielsen2020; @Peng2011] mainly among early career researchers [@McKiernan2016]. A key motivation for practicing openness is the notion that sharing all research artifacts will allow others to reproduce and verify results, thereby improving trust and verification of those findings. Jon Clarebout famously wrote in 1992 that reproducibility in computationally oriented fields had become so trivial that non experts could do the task. Clarebout's optimistic vision was a little ahead of its time as the tools and technologies available in the early nineties for literate programming presented too high of a barrier for most researchers to adopt without incurring a significant loss in productivity [@Barnes2010].

@@

```{r incentable-table, fig.cap = "Table of the mechanism for behavior change, the incentive, and our assessment of where the elements of Data, Code, and Computational Environment rank in terms of completing these aspects. We note that Data is often required, but the preceding steps are not required. This is in contrast to Code having no policy."}
knitr::include_graphics("figures/incentives_pyramid.png")
```

@@

Use a README file with your data project to communicate its purpose and contents. A README is often the first place users will look to learn more about a project [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards. Although data repositories contain a large number of datasets associated with peer reviewed papers, not all are usable (example: https://datadryad.org/stash/dataset/doi:10.5061/dryad.rn6k1).  Some of the more recent submissions contain usage notes and a README, which considerably improve reuse potential (e.g. https://datadryad.org/stash/dataset/doi:10.5061/dryad.3xsj3txbz). \

@@

Provide a license with your data. A license for data establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard one, these rules are unclear, and can lead to issues with use, attribution and citation. While there are a handful of licenses for data [@data_world_inc_common_2019], we recommend the CC0 license, as it provides simple rules for attribution, warranty, and use, and is very common. The CC0 allows data owners to waive all rights and data licensed as such can be freely shared, copied, modified, and distributed, even for commercial purposes without permission.  For more information on the benefit of using CC0, and why CCBY is not always appropriate, see https://osc.universityofcalifornia.edu/2016/09/cc-by-and-data-not-always-a-good-fit/. Short and full overviews of CCBY and CC0 can be found at [@cc0-short; @cc0-long].

@@

Make raw and analysis ready data available. Raw data is the first format of data provided before any cleaning has taken place (e.g., data in some binary or proprietary format). It should be placed in its own folder. Data used in analysis should be provided, especially if computationally intense to derive. It should be in its own folder, and to maximize interoperability, recorded in a plain text (e.g., csv, tsv) and if possible a binary format (e.g., .rds, .sav, .xlsx, Apache Parquet). The binary format allows for immediate reuse in the intended state, while the interoperability of plain text versions remain stable over time. Ideally data should be in "Tidy Data" form, where variables are in columns, observations in rows, and each cell has only one value [@Wickham2014].

@@

Share code used to transform data from its raw format into analysis ready form. This should be kept in the same place as raw data. Ideally, this transformation should involve only scripted languages (e.g, R, Python, Matlab, etc), but any non-scripted data preparation steps should be recorded in a plain text file. Authors should run the scripts in a clean environment to ensure everything runs and that the code doesn't rely on other local artifacts, such as their own folder layout not available to others.

@@

One critical advantage of depositing data in a trusted repository \@ref(avenues) is that it comes with many benefits such as preservation, indexing,  discoverability and most importantly, citation with persistent identifier. Authors can take an extra step by including their own CITATION file to give credit to a paper where the data and associated methods are described in detail.

@@

We'd like to thank the following people for helpful discussions and contributions to the paper as we were preparing it for publication: Anna Krystalli, Daniella Lowenberg, and Miles McBain. We'd also like to thank those who provided feedback on the paper when it went up for presubmission: Adrian Barnett, Raymond B Huey, Christian Thurn, Christina Maimone, and GitHub user, 'dschief001'. Finally, we thank the following organizations for their support ACEMS International Mobility Programme, the Helmsley Charitable Trust, the Gordon and Betty Moore Foundation, and the Sloan Foundation.

@@

Without data, there is no science. Science needs to be reproducible so we can trust the results, and progress as a field. But without data, we do not have true reproducibility. Currently, the role of data in reproducibility is not adequately discussed. We believe this is in part because discussions around data are centered on both data quality and data sharing. However, these are in fact two separate (but related) topics. We provide immediately usable guidelines to sharing data alongside your research, with minimal barriers, that a researcher could pick up and use today.

@@

Karthik is a senior data scientist at the Berkeley Institute of Data Science and a co-founder of the rOpenSci project. Karthik is also a senior PI at the Berkeley Initiative for Global Change Biology. Prior to joining Berkeley, he earned his PhD in Ecology & Evolution from the University of California, Davis. Karthik's interests are focused on reproducible research, as it applies to global change. Much of his recent work focuses on building tools and services around open data and growing diverse data science communities.
